The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2015-01-08 articles/2014/MWA-Radio-Environment/response-to-referee.txt

In the sentece: "alpha is typically -1.5 to -1.6"

Referee: "Presumably these nominal values of alpha are determined empirically?  This should be stated explicitly."

Answer: This is both empirically found as well as mathematically derived for uniform distributions of sources with a given average transmitter height. We think explaining this is a bit too much for this paper, so we have changed this sentence to "the exponent alpha is found by Offringa et al (2013b) to be typically -1.5 to -1.6."

"This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes..."

Referee: "Not sure "multi-variate" is the correct term here.  Perhaps non-stationarity and non-gaussanity of RFI is also a factor."

Answer:

Referee: "Such seemingly random impulsive RFI is ubiquitous - probably caused by ESD "sparking" and relay contacts."

Answer:

Figure 6.

"Are the centre channels of the PFB bands also flagged?  Otherwise, what are narrow horizontal lines?"

Answer: This is explained below the image: "[..] The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively."
These images are not correctly placed in the PASA-generated PDF, so this is a bit confusing.

### 2015-01-08 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Fixing ICRAR affil
Implementing changes requested by referee

\author[A.~R.~Offringa et al.]{A.~R.~Offringa\ANU$^,$\CAASTRO$^,$\ASTRON\thanks{Corresponding author. E-mail: \url{offringa@gmail.com}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~L.~Kaplan\UWisc,
N.~Barry\UW,
A.~P.~Beardsley\UW,
M.~E.~Bell\CASS$^,$\CAASTRO,
G.~Bernardi\SKASA$^,$\Rhodes$^,$\CfA,
J.~D.~Bowman\ASU, 
F.~Briggs\ANU$^,$\CAASTRO,
J.~R.~Callingham\USydney$^,$\CAASTRO,
R.~J.~Cappallo\Haystack, 
P.~Carroll\UW,
%A.~de~Oliveira-Costa
A.~A.~Deshpande\RRI, 
J.~S.~Dillon\MIT,
K.~S.~Dwarakanath\RRI,
A.~Ewall-Wice\MIT,
L.~Feng\MIT,
B.-Q.~For\UWA,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
L.~J.~Greenhill\CfA,
P.~Hancock\Curtin$^,$\CAASTRO,
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
D.~C.~Jacobs\ASU, 
M.~Johnston-Hollitt\Victoria,
A.~D.~Kapi\'{n}ska\UWA$^,$\CAASTRO,
H.-S.~Kim\UMelbourne$^,$\CAASTRO,
P.~Kittiwisit\ASU,
E.~Lenc\USydney$^,$\CAASTRO,
J.~Line\UMelbourne$^,$\CAASTRO,
A.~Loeb\CfA,
C.~J.~Lonsdale\Haystack, 
B.~McKinley\ANU$^,$\CAASTRO$^,$\UMelbourne,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
J.~Morgan\Curtin,
A.~R.~Neben\MIT,
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
S.~Paul\RRI,
B.~Pindor\UMelbourne,
J.~C.~Pober\UW,
T.~Prabu\RRI, 
P.~Procopio\UMelbourne$^,$\CAASTRO,
J.~Riding\UMelbourne$^,$\CAASTRO,
N.~Udaya~Shankar\RRI, 
S.~Sethi\RRI,
K.~S.~Srivani\RRI,
L.~Staveley-Smith\UWA$^,$\CAASTRO,
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
I.~S.~Sullivan\UW,
M.~Tegmark\MIT,
N.~Thyagarajan\ASU,
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
C.~M.~Trott\Curtin$^,$\CAASTRO,
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT,
C.~Wu\UWA,
J.~S.~Wyithe\UMelbourne$^,$\CAASTRO,
Q. Zheng\Victoria
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
%\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
\ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\SKASA{}SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
\Rhodes{}Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, University of Sydney, NSW 2006, Australia\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\UWA{}ICRAR, University of Western Australia, Crawley, WA 6009, Perth, Australia\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

@@

While there has been some success in RFI mitigation by actual removal of the interference while retaining the underlying data \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection and flagging of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg-2006,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequences of this approach are that a certain fraction of data is lost due to interference, and frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimize the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in processing the data from any interferometer. It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); \textsc{pieflag} \citep{pieflag-middelberg-2006} and \textsc{mirflag} \citep{lenc-mirflag-2010} mostly used for the Australia Telescope Compact Array (ATCA); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. \textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}. Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view. Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency axes. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but it was necessary to apply this correction for the data used here.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimising reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For data processing with the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995}, \textsc{wsclean} \citep{offringa-wsclean-2014} and/or the Fast Holographic Deconvolution software (\textsc{fhd}, \citealt{fhd-sullivan-2012}), the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output.

>>>>

\author[A.~R.~Offringa et al.]{A.~R.~Offringa\ANU$^,$\CAASTRO$^,$\ASTRON\thanks{Corresponding author. E-mail: \url{offringa@gmail.com}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~L.~Kaplan\UWisc,
N.~Barry\UW,
A.~P.~Beardsley\UW,
M.~E.~Bell\CASS$^,$\CAASTRO,
G.~Bernardi\SKASA$^,$\Rhodes$^,$\CfA,
J.~D.~Bowman\ASU, 
F.~Briggs\ANU$^,$\CAASTRO,
J.~R.~Callingham\USydney$^,$\CAASTRO,
R.~J.~Cappallo\Haystack, 
P.~Carroll\UW,
%A.~de~Oliveira-Costa
A.~A.~Deshpande\RRI, 
J.~S.~Dillon\MIT,
K.~S.~Dwarakanath\RRI,
A.~Ewall-Wice\MIT,
L.~Feng\MIT,
B.-Q.~For\UWA,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
L.~J.~Greenhill\CfA,
P.~Hancock\Curtin$^,$\CAASTRO,
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
D.~C.~Jacobs\ASU, 
M.~Johnston-Hollitt\Victoria,
A.~D.~Kapi\'{n}ska\UWA$^,$\CAASTRO,
H.-S.~Kim\UMelbourne$^,$\CAASTRO,
P.~Kittiwisit\ASU,
E.~Lenc\USydney$^,$\CAASTRO,
J.~Line\UMelbourne$^,$\CAASTRO,
A.~Loeb\CfA,
C.~J.~Lonsdale\Haystack, 
B.~McKinley\ANU$^,$\CAASTRO$^,$\UMelbourne,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
J.~Morgan\Curtin,
A.~R.~Neben\MIT,
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
S.~Paul\RRI,
B.~Pindor\UMelbourne,
J.~C.~Pober\UW,
T.~Prabu\RRI, 
P.~Procopio\UMelbourne$^,$\CAASTRO,
J.~Riding\UMelbourne$^,$\CAASTRO,
N.~Udaya~Shankar\RRI, 
S.~Sethi\RRI,
K.~S.~Srivani\RRI,
L.~Staveley-Smith\UWA$^,$\CAASTRO,
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
I.~S.~Sullivan\UW,
M.~Tegmark\MIT,
N.~Thyagarajan\ASU,
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
C.~M.~Trott\Curtin$^,$\CAASTRO,
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT,
C.~Wu\UWA,
J.~S.~Wyithe\UMelbourne$^,$\CAASTRO,
Q. Zheng\Victoria
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
%\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
\ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\SKASA{}SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
\Rhodes{}Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, University of Sydney, NSW 2006, Australia\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\UWA{}International Centre for Radio Astronomy Research, University of Western Australia, Crawley, WA 6009, Perth, Australia\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

@@

While there has been some success in RFI mitigation by actual removal of the interference while retaining the underlying data \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection and flagging of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg-2006,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequences of this approach are that a certain fraction of data is lost due to interference, and frequency \editmark{channels} that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimize the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each \editmark{tile comprising of a 4x4 array of dual-polarization dipoles}, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference \editmark{environment} of the MWA \editmark{may} improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose \editmark{unsurmountable} issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} conclude that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in processing the data from any interferometer. \editmark{One measure of the performance of an RFI detection method is its accuracy, which is often quantified by the average number of false-positive and true-positive detections resulting from the method.} It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple \editmark{amplitude} thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); \textsc{pieflag} \citep{pieflag-middelberg-2006} and \textsc{mirflag} \citep{lenc-mirflag-2010} mostly used for the Australia Telescope Compact Array (ATCA); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. \textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}. Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection \editmark{accuracy} for different band-pass shapes, time and frequency resolutions, and fields of view. Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency \editmark{domains}. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth \editmark{into 48 equal sub-bands and dividing each sub-band} by its Winsorized standard deviation \editmark{(see \citet{variance-estimates} for an explanation of Winsorized statistics}). This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only used during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but it was necessary to apply this correction for the data used here.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline \editmark{such that intermediate data can be kept in memory. This minimizes the reading and writing} of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter}\footnote{\editmark{Etymology: Cotter is a geographical area around Cotter River, near Mount Stromlo Observatory in Canberra.}} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally \editmark{ignored}, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the \editmark{visibilities} in frequency and/or time to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for \editmark{each} snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For data processing with the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995}, \textsc{wsclean} \citep{offringa-wsclean-2014} and/or the Fast Holographic Deconvolution software (\textsc{fhd}, \citealt{fhd-sullivan-2012}), the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics \editmark{calculated} over large time and frequency ranges. Therefore, \editmark{detection} accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without \editmark{increasing disk I/O overhead} it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation \editmark{into a number of shorter time segments and flag these independently}. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable \editmark{consequence} that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output \editmark{format}.

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most MWA observations are recorded at a resolution of 0.5~s and 40~kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not cause any significant time or bandwidth decorrelation up to the first beam null for MWA data. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$-values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation. \editmark{The central time/frequency of an output sample is set to the time/frequency mean of the corresponding input samples, independently of what input samples are flagged.}
